Prey movement shapes the development of predator expertise in a virtual bi-trophic system

### 2024-06-04 manuscript/manuscript.Rmd
updated intro

---
title: "Experience shapes individual foraging specialization and success in a virtual predator-prey system"
output: 
  word_document:
    reference_docx: manuscript-styles.docx
    fig_caption: yes
bibliography: references.bib
csl: chicago-author-date.csl
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(data.table)
library(brms)
library(flextable)
library(officer)
library(dplyr)

```{r}

path <- file.path(here::here(), "data")

dat <- fread(
  file.path(path, "FraserFrancoetal2023-data.csv"),
  select = c("timestamp", "match_id", "realm_id",
             "environment_id", "avatar_id", "predator_id",
             "game_duration", "total_xp_pred", "cumul_xp_pred",
             "hunting_success", "pred_speed", "prey_avg_speed"))

@@

# To have a date column
dat[, date := as.POSIXct(sub(" .*", "", timestamp))]

@@

The capacity of predators to match their tactic to their prey is expected to drive the outcome of predator-prey interactions. To do so, predators need to practice their tactics extensively to develop the skills to be successful hunters, which may result in individual foraging specialization. Yet, there are limited empirical assessments showing links between experience, foraging specialization, and hunting success at the individual level, due to the challenges of monitoring direct interactions in the wild. Here, we used a virtual predator-prey system (the game *Dead by Daylight*) to investigate how experience shapes individual foraging specialization and success in predators across repeated interactions with their prey. We show that predators optimized prey consumption as they gained experience. Prey speed was an important mediator of this relationship, resulting in differences among predators in the development of expertise. Throughout experience, 68% of predators became either increasingly specialized by always moving at a fast pace, or flexible by switching between slow and fast speeds. The predators’ strategies were partially matched to the preys’ speed, suggesting that changes in hunting behaviour were driven by repeated encounters with prey. Flexible and specialist foragers achieved similar success overall. Our findings suggest that experience may be mechanism promoting behavioural diversification in predator-prey systems.

Keywords: foraging behaviour, reciprocal behavioural plasticity, learning, antipredator behaviour, virtual ecology, Dead by Daylight

@@

Predators express a diverse array of foraging tactics to capture their prey. In natural populations, differences among individuals in tactics such as ambushing and fast movement speeds are associated with differences in prey consumption [@toscano2014;@patrick.weimerskirch2014a;@toscano.etal2016]. Prey usually respond to predation cues by reducing their activity, or use rapid or unpredictable movements to successfully escape when they are chased [@walker.etal2005;@kelley.magurran2011;@herbert-read.etal2017]. This results in reciprocal interactions where predators fine-tune their tactics to their prey, suggesting that they learn and become efficient hunters through repeated encounters [@estes.etal2003;@woo.etal2008;@phillips.etal2017]. These behavioural adjustments with experience can shape the outcome of predator-prey interactions, and may have larger implications for community and ecosystem dynamics [@wooster.etal2023].

Empirical studies on human and non-human hunters show that individuals optimize foraging efficiency (e.g. search and handling times, return rates) by learning associative images or prey and environmental cues [@edwards.jackson1994;@morse2000;@macdonald2007;@reid.etal2010;@wilson-rankin2015]. As predators practice and refine their tactics, they acquire expertise (e.g. skills, knowledge) that allows them to outperform novice hunters [@dukas2019;@wooster.etal2023]. Through this process, it is hypothesized that individuals will gradually specialize in a foraging tactic (e.g. always moving fast) if it allows them to maintain successful prey captures at each attempt [@tinker.etal2009;@dukas2019]. This should occur when predators learn to hunt in environments where resources are predictable (or stable) because they require less energy and time to search for, capture, and handle prey [@weimerskirch2007;@woo.etal2008;@potier.etal2015]. Such conditions favor specialist hunters because there would be higher costs to switch tactics for the same type of prey. An alternative hypothesis is that learning is a mechanism driving behavioural flexibility. For instance, predators can gain expertise and information on their prey through repeated hunting atempts, and may hunt optimally by learning to adjust their tactic to the type of prey that they encounter [@stephens1993;@ishii.shimada2010;@kelley.magurran2011]. While there are costs to attempt novel tactics by trial and error in stable environments, predators that hunt in environments under fluctuating resources can reduce the consequences of uncertainty by learning to switch hunting tactics [@dall2010;@mathot.etal2012]. Nonetheless, complex feedback between ecological interactions and learning make it challenging to predict under which circumstances predators should specialize with experience [@stephens1993;@ishii.shimada2010;@vilapouca2021;@beyts2023].

There is increasing evidence suggesting that the emergence of behavioural specialization and its fitness consequences are driven by resource fluctuations [@patrick.weimerskirch2014b;@vandenbosch.etal2019;@santoro.etal2019;@manlick.etal2021]. For example, @szopa-comley.ioannou2022 showed that predatory blue acaras (*Andinoacara pulcher*) reduce their maximum approach speed when repeatedly exposed to prey that escape unpredictably, enabling them to be as successful as those exposed to predictable prey. Prey that use fast escapes are also proven to be more difficult to hunt [@walker.etal2005;@kelley.magurran2011;@martin.etal2022], and predators that exclusively encounter such prey should benefit from learning to specialize in fast pursuits. If predators that hunt in different environments adjust their tactics to local prey fluctuations, then specialist and flexible hunters can both attain similar hunting success and coexist [@phillips.etal2017]. Therefore, empiricists should investigate the link between foraging specialization and success across time and contexts.

A recurring challenge impeding research on predator-prey behavioural interactions is the need to collect data simultaneously on both the predator and the prey. Similar to agent-based simulations, online videogames can be useful systems to mitigate these challenges and test specific ecological hypotheses. They provide controlled virtual environments with the advantage of having real players that interact in the virtual space. For example, in the predator-prey videogame *Dead by Daylight*, four prey players need to forage for resources while avoiding predation by a fifth player. In this virtual system, the predator population is composed of individuals that either ambush or hunt at high speeds (i.e. mean movement speed along a slow-fast continuum), and their success is driven by the movement of the prey [@fraserfranco.etal2022]. We found that prey can increase their chances of survival by cooperating and moving fast when they forage [@cere.etal2021;@fraserfranco.etal2022, Santostefano et al. in press]. Of course, the life of neither the predator nor the prey players are at stake, such that emerging patterns could be driven by their motivation to win and not "true" survival. Yet, the game ellicits natural reactions in players such as fear from predation (personal observations), which corroborates with another virtual ecological study showing that predation drives individual variation in risk perception [@beauchamp2020]. In a different system, @lymbery.etal2023 showed that complex environments favor strong soldiers over large armies in a virtual game as well as in the laboratory with two ant species. These observations outline how ecological patterns can emerge from interactions in virtual systems with fixed rules. Videogames also generate large volumes of data on interacting players throughout their lifetime in the game under realistic, controlled, and repeatable ecological scenarios. Hence, virtual systems such as *Dead by Daylight* offer the opportunity to tackle fundamental questions about the role of experience on predator-prey interactions.

In this study, we assess how repeated experience shapes predator foraging specialization and success using data from players in *Dead by Daylight*. Following @fraserfranco.etal2022 we use the predators' mean movement speed along a slow-fast continuum as a proxy for the foraging tactic (ambush-cursorial), and define the level of individual foraging specialization as the intra-individual variance (i.e. IIV) in movement speed across matches, where low IIV translates to individual foraging specialization (e.g. always using a cursorial tactic). First, we investigate how individual hunting success changes with experience (i.e. how individuals develop their hunting expertise). We hypothesize that the predators' success will increase with experience up to a certain level where it will stagnate. We expect this pattern to vary among individuals due to differences in prey encounters. Second, we assess how foraging specialization emerges with experience. If all predators encounter similar groups of prey, we predict that individuals (and thus the population) should all specialize in similar speeds. In contrast, if all predators encounter varying groups of prey, then they should all converge towards flexible speeds. In both scenarios, differences among individuals in IIV across experience should be low (i.e. similar individual foraging specialization), whereas the population variance would either decrease (specialization) or increase (flexibility). Alternatively, differences among individuals in foraging specialization may emerge if they experience different interactions with their prey. In this case, we expect predators that encountered similar groups of prey across experience to specialize in similar speeds, while predators that encountered heterogeneous groups across matches should adopt flexible movement speeds, resulting in an increase in among individual differences in IIV with experience. If we detect such prey-dependent fine-tuning with experience, then all hunters along the flexible-specialist continuum should attain equal success.

@@

Our population consists of `r toString(format(length(unique(dat$predator_id))))` predator players with a total record of `r toString(formatC(length(unique(dat$match_id)), big.mark = " "))` matches. The predator-players' experience varied between `r toString(format(min(dat$total_xp_pred)))` and `r toString(format(max(dat$total_xp_pred)))` matches played. These matches lasted between `r toString(format(round(min(dat$game_duration / 60))), digits = 2)` and `r toString(format(round(max(dat$game_duration / 60))), digits = 2)` min (mean = `r toString(format(round(mean(dat$game_duration / 60))), digits = 2)` min). The following information is collected and reported for every match : the player's anonymous ID, its avatar (i.e. the predator character chosen with its specific skill-gameplay mechanics), the game environment, the predator-player's experience along with its speed, the mean speed of the groups of prey the player encountered, and their mean rank (a proxy for prey skill). The ranking system in *DBD* was implemented by the company to pair players in a match based on their skill (https://deadbydaylight.fandom.com/wiki/Rank), and failing to account for it would prevent us from detecting a change in the predator's foraging success with experience.

We analyzed the predator's mean speed and the mean speed of the prey group encountered by the predator. The predator's mean speed is measured as the mean distance traveled per second during a match (mean= `r toString(format(round(mean(dat$pred_speed), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$pred_speed), digits = 2), nsmall = 2))` m/s). We measured the preys' speed as the mean travel speed of the four individual prey in a match (mean = `r toString(format(round(mean(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` m/s). We defined hunting success as the number of prey consumed during the match (min = `r toString(format(min(dat$hunting_success)))`, max = `r toString(format(max(dat$hunting_success)))`). Lastly, we defined the predator's cumulative experience as the number of matches played as the predator prior to the match being monitored. For example, the first match of a player would have a cumulative experience value of 0, while the tenth match would have a value of 9. We did not account for matches where predators played as the prey. We also categorized predators for each match based on their cumulated experience. We labeled predators as novices for matches where they had cumulated less than 100 matches, intermediate for matches where they had cumulated between 100 and 299 matches, and advanced for matches where they had cumulated more than 299 matches (max 499). Since our goal was to monitor predator players throughout their experience and that they all played at least 300 matches, they all appeared in each of the three experience categories.

@@

### Development of expertise

@@

### Foraging behaviour and predator-prey interactions across experience

We tested whether foraging behaviour changed with experience, and compared the success of predators along the flexible-specialist hunter continuum, using a multivariate double-hierarchical generalized linear model (MDHGLM) [@lee.nelder2006;@cleasby.etal2015;@mitchell.etal2016]. For each experience level, we fitted the model on three traits: the predator's speed ($Y_{1}$) and the prey's speed ($Y_{2}$) using a Gaussian error distribution, and the predator's hunting success ($Y_{3}$) using a beta-binomial error distribution. Fixed and random effect structure were trait-specific and the same for all experience levels. The predator and the prey speed included a mean and a dispersion part (eqn 1.1, 1.2 and 2.1, 2.2). For both traits, we controlled for the standardized prey rank on both the mean and the dispersion parts of the model. We included game environment, predator avatar, and individual predator identity random effects on the mean part of the model. The dispersion part included individual predator identity random effects exclusively. For hunting success, we modeled the mean part only (eqn 3) and controlled for the standardized prey rank and game duration, and included an individual identity random effect.

The model simultaneously estimated five among-individual variance components with their covariances for each experience level, resulting in a 15x15 (co)variance matrix. For each experience level, the model estimated among individual variance in mean predator speed, mean prey speed, and mean hunting success. This allows us to test for individual differences in tactic use (i.e. along the slow-fast continuum), in the mean speed of the prey encountered, and in mean prey consumption. The model also estimated among individual variance in intra-individual variation (IIV) for both prey and predator speed. This allows us to test if predators do not experience the same degree of prey heterogeneity and if they differ in individual specialization. Finally, the model estimated all pairwise covariances at the individual level among traits, among experience, and across traits and experience. The resulting 15x15 (co)variance matrix thus provides the structure of the predator-prey trait interactions as well as the relationship between specialization and success across experience at the individual-level.

The model can be written as:

```{r model_equations, ft.align = "left"}

# Prepare equations
#eqs <- c(
#  "Y_{1}(T_{j}) = X_{1}(T_{j})\\ b_{1}(T_{j}) + M_{1}(T_{j})\\ en_{1}(T_{j}) + N_{1}(T_{j})\\ av_{1}(T_{j}) + Z_{1}(T_{j})\\ #id_{1}(T_{j}) + e_{1}(T_{j})",
#  "log(\\sigma_{e_{Y_{1}(T_{j})}}) = X_{d1}(T_{j})\\ b_{d1}(T_{j}) + Z_{d1}(T_{j})\\ id_{d1}(T_{j})",
#  "Y_{2}(T_{j}) = X_{2}(T_{j})\\ b_{2}(T_{j}) + M_{2}(T_{j})\\ en_{2}(T_{j}) + N_{2}(T_{j})\\ av_{2}(T_{j}) + Z_{2}(T_{j})\\ #id_{2}(T_{j}) + e_{2}(T_{j})",
#  "log(\\sigma_{e_{Y_{2}(T_{j})}}) = X_{d2}(T_{j})\\ b_{d2}(T_{j}) + Z_{d2}(T_{j})\\ id_{d2}(T_{j})",
#  "Y_{3}(T_{j}) = X_{3}(T_{j})\\ b_{3}(T_{j}) + Z_{3}(T_{j})\\ id_{3}(T_{j}) + e_{3}(T_{j})"
#)

#eqs <- c(
#  "Y_{1}(T_{j}) = T^2_{j}(X_{1}\\ b_{1}) + T^2_{j}(M_{1}\\ en_{1}) + T^2_{j}(N_{1}\\ av_{1}) + T^2_{j}(Z_{1}\\ id_{1}) + e_{1}(T_{j})",
#  "log(\\sigma_{e_{Y_{1}(T_{j})}}) = T^2_{j}(X_{d1}\\ b_{d1}) + T^2_{j}(Z_{d1}\\ id_{d1})",
#  "Y_{2}(T_{j}) = T^2_{j}(X_{2}\\ b_{2}) + T^2_{j}(M_{2}\\ en_{2}) + T^2_{j}(N_{2}\\ av_{2}) + T^2_{j}(Z_{2}\\ id_{2}) + e_{2}(T_{j})",
#  "log(\\sigma_{e_{Y_{2}(T_{j})}}) = T^2_{j}(X_{d2}\\ b_{d2}) + T^2_{j}(Z_{d2}\\ id_{d2})",
#  "Y_{3}(T_{j}) = T^2_{j}(X_{3}\\ b_{3}) + T^2_{j}(Z_{3}\\ id_{3}) + e_{3}(T_{j})"
#)

eqs <- c(
  "Y_{1j} = X_{1j}\\ b_{1j} + M_{1j}\\ en_{1j} + N_{1j}\\ av_{1j} + Z_{1j}\\ id_{1j} + e_{1j}",
  "log(\\sigma_{e_{Y_{1j}}}) = X_{d1j}\\ b_{d1j} + Z_{d1j}\\ id_{d1j}",
  "Y_{2j} = X_{2j}\\ b_{2j} + M_{2j}\\ en_{2j} + N_{2j}\\ av_{2j} + Z_{2j}\\ id_{2j} + e_{2j}",
  "log(\\sigma_{e_{Y_{2j}}}) = X_{d2j}\\ b_{d2j} + Z_{d2j}\\ id_{d2j}",
  "Y_{3j} = X_{3j}\\ b_{3j} + Z_{3j}\\ id_{3j} + e_{3j}"
)

eqs_num <- c(
    "(eqn. 1.1)",
    "(eqn. 1.2)",
    "(eqn. 2.1)",
    "(eqn. 2.2)",
    "(eqn. 3)"
)

# Combine into a dataframe
df <- data.frame(
    formula = eqs,
    number = eqs_num
)

# Create a table to display on Word
ft <- flextable(df)

# Transform values as a mathematical formula
ft <- compose(
  x = ft, j = 1,
  value = as_paragraph(
    as_equation(formula, height = .5)
  )
)

# Customize table
my_theme <- function(x, ...) {
  x <- colformat_double(x, big.mark = " ",
                        decimal.mark = ".",
                        digits = 2)
  x <- border_remove(x)
  x <- delete_part(x, part = "header")
  x <- align(x, j = 1, align = "left", part = "body")
  x <- align(x, j = 2, align = "right", part = "body")
  x <- font(x, fontname = "Times New Roman", part = "body")
  x <- fontsize(x, size = 12, part = "body")
  x <- width(x, j = 1, width = 13, unit = "cm")
  x <- width(x, j = 2, width = 2.26, unit = "cm")
}

# Apply theme and print
ft <- my_theme(x = ft)
ft

where $j$ is the index of the three experience levels (i.e. novice, intermediate, advanced). The $d$ subscript indicates that a term is from the dispersion part of the model. The $b$ terms are vectors of fixed effects with their design matrices $X$ specific to each trait and experience level. The terms $en$, $av$, and $id$ are the vectors of random environment, random avatar, and random individual identity effects associated with their design matrices $M$, $N$, and $Z$ specific to each trait and experience level. For all experience levels of $Y_{1}$ and $Y_{2}$, we assumed that the game environment and avatar random effects followed a Gaussian distribution with estimated standard deviation ($en_{12j} \sim N(0,\ I\ \sigma_{en_{12j}}^{2})$ and $av_{12j} \sim N(0,\ I\ \sigma_{av_{12j}}^{2})$). $I$ is the identity matrix, with $I\sigma^{2}$ indicating that the random effects are independently and identically distributed. In addition, we assumed that the residuals follow a Gaussian distribution with observation-specific variance $e_{12j} \sim N(0,\ Diag\{\sigma_{e_{12j}}^{2}\})$. The 15 individual identity random effects $id_{ij}$ across all response variables and experience levels (five for each of the three experience levels) are indexed in vector $\alpha$. The $\alpha$ vector follows a multivariate Gaussian distribution $\alpha \sim MVN(0,\ \Sigma)$ where $\Sigma$ is a 15x15 (co)variance matrix of among individual variation across all response variables and experience levels, with $\sigma_{id_{ij}}^{2}$ on the diagonal and covariances off-diagonal.

We used weakly informative Gaussian priors for the prey rank ($N(0,\ 1^{2})$: eqn. 1.1 to 2.2) and game duration ($N(0.5,\ 0.5^{2})$: eqn. 3). Based on previous results showing that the mean predator speed revolves around 3 m/s [@fraserfranco.etal2022], we applied a weakly informative Gaussian prior on the intercept for predator and prey speed ($N(3,\ 0.5^{2})$: eqn. 1.1 and 2.1). Similarly, we used a weakly informative Gaussian prior on the intercept for the dispersion part ($N(0,\ 1^{2})$: eqn 1.2 and 2.2). For hunting success, we applied a positive Gaussian prior on the precision parameter ($N(2,\ 1^{2})$: eqn. 3) and a negative Gaussian prior ($N(-4,\ 1^{2})$: eqn 3) on the intercept because hunting success should be close to 0 at the lowest game duration. For every predicted variable, we applied a weakly informative half-Gaussian prior ($N(0,\ 1^{2})$) on the random effects for the mean part and the dispersion part of the model. We used a regularizing $LKJ(2)$ prior on the correlation matrix.

After inspection of the results, we found that key parameter values for our hypotheses were in the direction of our predictions, but they were largely driven by the very slow speed of one predator. Removing this individual reduced the estimates by a significant margin. To remain conservative, we thus present the results of the model without this individual in the main text, and provide the results of the model including it in the Appendix S3.

@@

```{r}

# Open percentages table
path <- file.path(here::here("outputs", "03_outputs_model-processing"))
percent_tab1 <- data.table(readRDS(file.path(path, "GAMM_percentages.rds")))
percent_tab2 <- data.table(readRDS(file.path(path, "DHMLM_percentages.rds")))

@@

# Open the loo table
path <- file.path(here::here("outputs", "02_outputs_model-validation"))
loo_tab <- data.table(readRDS(file.path(path, "GAMM_loo-cv-table.rds")))

@@

# Custom header
my_header <- data.frame(
   col_keys = c("model",
                "elpd_diff",
                "se_diff",
                "elpd_loo",
                "se_elpd_loo"),
   line1 = c("model",
             "elpd \ndifference",
             "sd \ndifference",
             "elpd loo \nvalue",
             "elpd loo \nstandard error"),
   stringsAsFactors = FALSE
)

# Custom theme
my_theme <- function(x, ...) {
   x <- colformat_double(x, big.mark = " ",
                         decimal.mark = ".",
                         digits = 2)
   x <- set_table_properties(x, layout = "autofit")
   x <- border_remove(x)
   std_border <- fp_border(width = 1, color = "black")
   x <- hline_top(x, part = "all", border = std_border)
   x <- hline_bottom(x, part = "all", border = std_border)
}

@@

knitr::include_graphics(
  here::here("outputs", "04_outputs_figures", "figure1.png")
)

## Changes in foraging behaviour, prey behaviour, and hunting success

The mean speed of the predator population remained stable with experience (Figure 2 and Appendix S1: Table S1). Likewise, the variation in speed did not change with experience, indicating that foraging specialization remained stable at the population level (Figure 2 and Appendix S1: Table S1). At the individual level, predators differed slightly in their mean speed across experience levels, indicating marginal differences in hunting tactics (Figure 2 and Appendix S1: Table S2). However, they displayed important differences in foraging specialization (Appendix S1: Table S2) which also increased slightly with experience as there was a 0.11 unit increase in among individual differences in IIV from novice to advanced (Figure 2). `r toString(scales::percent(percent_tab2[1,]$percentage, accuracy = 1))` of the population switched from a flexible to a specialized hunting tactic, and vice-versa, as predators gained experience (i.e. >0.2 change in standard deviation; Figure 3). In contrast, `r toString(scales::percent(percent_tab2[2,]$percentage, accuracy = 1))` displayed lower changes (i.e. >0.05 and <0.2 change in standard deviation) and `r toString(scales::percent(percent_tab2[3,]$percentage, accuracy = 1))` showed almost no change with experience (i.e. <0.05 change in standard deviation).

```{r fig2, out.width = "100%", fig.cap = "**Figure 2**. Median posterior differences and HPD intervals comparing predator foraging behaviour at the population and individual level across experience stages. The behavioural parameter predicted by the MDHGLM being compared is on the y axis, and the difference in parameter values between two experience stages is on the x axis. The pairwise comparisons between experience stages are displayed across the three panels A, B, and C. We compare the mean speed of the predator population, the speed of the prey it encountered, and its hunting success across experience (population mean). We also compare the population's foraging specialization and variation in speed of the prey it encountered across experience (population variance). At the individual level, we compare among individual differences in speed, the speed of the prey encountered, and hunting success (individual variation mean). Lastly, we compare individual variation in IIV across experience, indicating whether individual differences in foraging specialization and in the variation of the groups of prey encountered changed with experience (individual variation IIV)."}

knitr::include_graphics(
  here::here("outputs", "04_outputs_figures", "figure2.png")
)

As the predator population gained experience, there was strong evidence that the prey increased their speed, although slightly (Figure 2 and Appendix S1: Table S1). The population also encountered groups of prey using different speeds through time (Figure 2 and Appendix S1: Table S1). At the individual level, all predators encountered prey with similar average speeds throughout experience (Figure 2 and Appendix S1: Table S2). Yet, even if they experienced similar average prey speeds across all groups encountered, there were important differences among predators in how (dis)similar were the groups they encountered (Appendix S1: Table S2). Thus, some predators encountered similar groups (i.e. all groups using similar speeds) while others encountered distinct groups (i.e. some slower and some faster). These differences among individuals in IIV for prey encounters increased only marginally with experience (Figure 2).

Similar to the GAMMs, the predator population's mean hunting success increased with experience (Figure 2 and Appendix S1: Table S1), with mean prey consumption revolving around two prey. There were also important differences among individuals in mean hunting success (Appendix S1: Table S2), but they remained stable across experience (Figure 2).

```{r fig3, out.width = "100%", fig.cap = "**Figure 3**. Individual behavioural distribution of the predators' speed drawn from the MDHGLM estimates for individuals that had (A) the greatest increase in specialization and (B) the greatest increase in flexibility. The background (gray) distributions are for when individuals were novice hunters, and the foreground (blue) distributions with solid contour lines are for when they were advanced hunters. Both figure panels are ordered by ascending degree of increase in either specialization or flexibility. We built the figure by first subtracting the estimated standard deviation of all individuals as novices with their standard deviation as advanced hunters, and then selected those with the greatest difference using the 25% and 75% quantiles. Individuals on panel A have an increase in specialization equal or greater than 0.2 standard deviations, while those on panel B have an increase in flexibility equal to or greater than 0.2 standard deviations."}

knitr::include_graphics(
  here::here("outputs", "04_outputs_figures", "figure3.png")
)

## Predator-prey interactions across experience

Predators that were on average faster had lower IIV in speed (Figure 4A-B and Appendix S4: Table S1). Thus, faster predators specialized on fast movement, while slower predators were more flexible in their movement. As predators gained experience, these strategies were increasingly defined as the correlation changed by 0.16 units, from -0.62 (-0.91, -0.41 $95\%\ CI$) for when predators were novices to -0.78 (-1.00, -0.53 $95\%\ CI$) as they reached the advanced stage (Figure 4C). Predators that were on average faster also tended to encounter groups of prey that were on average faster and more similar (Figure 4A-B and Appendix S4: Table S1). There was however no evidence that these correlations increased with experience (Figure 4C)

There was no correlation (-0.01 (-0.17, 0.09 $95\%\ CI$)) between mean prey speed and IIV in predator speed when predators were novice (Figure 4A), but the correlation increased to -0.16 (-0.33, -0.04 $95\%\ CI$) at the advanced stage (Figure 4B), such that advanced hunters that encountered faster prey tended to be more specialized (Figure 4C). The predators' IIV in speed and IIV in prey speed were weakly positively correlated (Figure 4A-B and Appendix S4: Table S1). Hence, predators that encountered groups of prey with similar speeds tended to be more specialized in their tactic, while those that encountered groups of prey with varying speeds tended to be more flexible. We did not find evidence that this correlation increased with experience (Figure 4C).

>>>>

---
title: "Prey movement shapes the development of predator expertise in a virtual bi-trophic system"
output: 
  word_document:
    reference_docx: manuscript-styles.docx
    fig_caption: yes
bibliography: references.bib
csl: chicago-author-date.csl
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(data.table)
library(brms)
library(flextable)
library(officer)
library(dplyr)
```

```{r}
path <- file.path(here::here(), "data")

dat <- fread(
  file.path(
    path, "FraserFrancoetal2023-data.csv"
  ),
  select = c(
    "timestamp", "match_id", "realm_id",
    "environment_id", "avatar_id", "predator_id",
    "game_duration", "total_xp_pred", "cumul_xp_pred",
    "hunting_success", "pred_speed", "prey_avg_speed"
  )
)

@@

# To have a date column
dat[, date := as.POSIXct(sub(" .*", "", timestamp))]
```

@@

The capacity of predators to match their hunting tactic to their prey is expected to drive the outcome of predator-prey interactions. To do so, predators need to practice their tactics extensively to develop the expertise to be successful hunters. Yet, there are limited empirical assessments showing links between experience and hunting success at the individual and population levels due to the challenges of monitoring direct interactions in the wild. Here, we use a virtual predator-prey system (the game *Dead by Daylight*) to investigate how experience shapes individual and population hunting success in predators across repeated interactions with their prey. We show that predators optimized prey consumption as they gained experience. Prey speed was an important mediator of this relationship, resulting in differences among predators in the development of expertise. **Décrire le reste des résultats...**

Keywords: foraging behaviour, experience, learning, antipredator behaviour, virtual ecology, Dead by Daylight

@@

The ecological consequences of predation are defined by the sum of consumptive and nonconsumptive effects (CEs and NCEs) during predator-prey interactions [refs]. Recent advances in our understanding of the behavioural mechanisms (e.g. reciprocal plasticity, foraging specialization) driving prey consumption at the individual and population level have allowed ecologists to rely on foraging strategies to predict the outcome of (N)CEs and their effect on prey population dynamics [refs]. For example, empirical studies show that prey tend to increase vigilance at the expense of foraging in the presence of ambush predators (NCEs) [@preisser2007]. Because ambush predators are opportunistic hunters, they also tend to consume a greater range of prey types (CEs) compared to cursorial predators [@glaudas2019].

While the foraging mode of predators is a reliable predictor of prey consumption, predators need to acquire expertise by practicing and learning the proper skills to successfuly capture their prey [@dukas2019;wooster.etal2023]. The aquisition of expertise is predicted to be paramount in many predator systems where hunting tactics are learned (e.g. orca whales). Yet, there is only fraction of evidence showing the link between experience and success in natural predators, impeding our ability to accurately predict the consequences of predation on prey populations.

Empirical studies on human and non-human hunters show that individuals optimize foraging efficiency (e.g. search and handling times, return rates) by learning associative images or using cues from their prey and their environment [@edwards.jackson1994;@morse2000;@macdonald2007;@reid.etal2010;@wilson-rankin2015]. Predators can integrate this information to use the appropriate hunting tactic (e.g. ambushing or fast movement speeds) for the type of prey that they encounter [@toscano2014;@patrick.weimerskirch2014a;@toscano.etal2016]. However, prey often use antipredator tactics such as rapid or unpredictable movements which are effective to successfully escape predators [@walker.etal2005;@kelley.magurran2011;@herbert-read.etal2017;@szopa-comley.ioannou2022]. This results in reciprocal interactions where predators need to learn to predict the movement of their prey to capture them. Therefore, antipredator tactics may impede the acquisition of hunting expertise by interfering with learning processes and reducing the success of predators.

A recurring challenge impeding research on predator-prey behavioural interactions is the need to collect data simultaneously on both the predator and the prey. Similar to agent-based simulations, online videogames can be useful systems to mitigate these challenges and test specific ecological hypotheses. They provide controlled virtual environments to test ecological hypotheses [see @lymbery.etal2023] with the advantage of having real players that interact in the virtual space. For example, in the predator-prey videogame *Dead by Daylight*, four prey players need to forage for resources while avoiding predation by a fifth player. In this virtual system, the predator population is composed of individuals that either ambush or hunt at high speeds (i.e. mean movement speed along a slow-fast continuum), and their success is driven by the movement of the prey [@fraserfranco.etal2022]. The prey can increase their chances of survival by cooperating and moving fast when they forage [@cere.etal2021;@fraserfranco.etal2022;@santostefano.etal2024]. The game also ellicits natural reactions in players such as fear from predation (personal observations), which corroborates with another virtual ecological study showing that predation drives individual variation in risk perception [@beauchamp2020]. These observations outline how ecological patterns can emerge from interactions in virtual systems with fixed rules. Videogames also generate large volumes of data on interacting players throughout their lifetime in the game under realistic, controlled, and repeatable ecological scenarios. Hence, virtual systems such as *Dead by Daylight* offer the opportunity to tackle fundamental questions about the role of experience on predator-prey interactions.

In this study, we assess how repeated experience shapes predator foraging success using data from players in *Dead by Daylight*. First, we investigate how individual hunting success changes with experience (i.e. how individuals develop their hunting expertise). We hypothesize that the predators' success will increase with experience up to a certain level where it will stagnate. We expect this pattern to vary among individuals due to differences in prey encounters. Second, we investigate how prey movement influences the development of expertise. If all individual predators encounter groups of prey with similar speeds, we predict that their gain in expertise will be similar.

In contrast, based on previous results showing that faster prey are more difficult to hunt, then...

if all predators encounter varying groups of prey, then they should all converge towards flexible speeds. In both scenarios, differences among individuals in IIV across experience should be low (i.e. similar individual foraging specialization), whereas the population variance would either decrease (specialization) or increase (flexibility).

@@

Our population consists of `r toString(format(length(unique(dat$predator_id))))` predator players with a total record of `r toString(formatC(length(unique(dat$match_id)), big.mark = " "))` matches. The predator-players' experience varied between `r toString(format(min(dat$total_xp_pred)))` and `r toString(format(max(dat$total_xp_pred)))` matches played. These matches lasted between `r toString(format(round(min(dat$game_duration / 60))), digits = 2)` and `r toString(format(round(max(dat$game_duration / 60))), digits = 2)` min (mean = `r toString(format(round(mean(dat$game_duration / 60))), digits = 2)` min). The following information is collected and reported for every match : the player's anonymous ID, its avatar (i.e. the predator character chosen with its specific skill-gameplay mechanics), the game environment, the predator-player's experience, the mean speed of the groups of prey the player encountered, and their mean rank (a proxy for prey skill). The ranking system in *DBD* was implemented by the company to pair players in a match based on their skill (https://deadbydaylight.fandom.com/wiki/Rank), and failing to account for it would prevent us from detecting a change in the predator's foraging success with experience.

We analyzed the mean speed of the prey group encountered by the predator. We measured the preys' speed as the mean travel speed of the four individual prey in a match (mean = `r toString(format(round(mean(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` m/s). We defined hunting success as the number of prey consumed during the match (min = `r toString(format(min(dat$hunting_success)))`, max = `r toString(format(max(dat$hunting_success)))`). Lastly, we defined the predator's cumulative experience as the number of matches played as the predator prior to the match being monitored. For example, the first match of a player would have a cumulative experience value of 0, while the tenth match would have a value of 9. We did not account for matches where predators played as the prey. We also categorized predators for each match based on their cumulated experience. We labeled predators as novices for matches where they had cumulated less than 100 matches, intermediate for matches where they had cumulated between 100 and 299 matches, and advanced for matches where they had cumulated more than 299 matches (max 499). Since our goal was to monitor predator players throughout their experience and that they all played at least 300 matches, they all appeared in each of the three experience categories.

@@

### Model specification

@@

@@

```{r}
# Open percentages table
path <- file.path(here::here("outputs", "outputs_model-processing"))
percent_tab1 <- data.table(readRDS(file.path(path, "GAMM_percentages.rds")))

@@

# Open the loo table
path <- file.path(here::here("outputs", "outputs_model-validation"))
loo_tab <- data.table(readRDS(file.path(path, "GAMM_loo-cv-table.rds")))

@@

# Custom header
my_header <- data.frame(
  col_keys = c(
    "model",
    "elpd_diff",
    "se_diff",
    "elpd_loo",
    "se_elpd_loo"
  ),
  line1 = c(
    "model",
    "elpd \ndifference",
    "sd \ndifference",
    "elpd loo \nvalue",
    "elpd loo \nstandard error"
  ),
  stringsAsFactors = FALSE
)

# Custom theme
my_theme <- function(x, ...) {
  x <- colformat_double(
    x, big.mark = " ",
    decimal.mark = ".",
    digits = 2
  )
  x <- set_table_properties(x, layout = "autofit")
  x <- border_remove(x)
  std_border <- fp_border(width = 1, color = "black")
  x <- hline_top(x, part = "all", border = std_border)
  x <- hline_bottom(x, part = "all", border = std_border)
}

@@

knitr::include_graphics(
  here::here("outputs", "outputs_figures", "figure1.png")
)
```
